Building a continuous benchmarking ecosystem in bioinformatics

### 2024-11-07 manifesto-main.tex
Mark's changes

Benchmarking (further defined below) is a critical step in method development and is typically published as either `methods-development papers' (MDPs) where new methods are compared against existing ones; or `benchmark-only papers' (BOPs), where a set of existing methods are compared in a more neutral way, as defined in \cite{Cao2023-jz}. A benchmarking study can be run by a single person or a small group, or it can be organized as a `challenge' or within a hackathon. In this work, we present a perspective on building a robust benchmarking ecosystem in bioinformatics, coming primarily from the methodologist's viewpoint (researchers developing computational tools).  While our focus is bioinformatics, we aim to abstract principles and connect to similar challenges in adjacent fields.

In our discussion, we outline concepts and specifications for an ideal benchmarking system; we examine philosophical issues like gatekeeping and trust; and provide some technical recommendations, such as standardization practices.

@@

Benchmarks are useful for a broad set of stakeholders. For a \emph{data analyst}, with an aim to perform a specific type of analysis on a specific dataset, benchmarks are helpful tools to find a suitable method for the task. Hence, it is important that benchmarks include datasets similar to the one of interest to the analyst, as method performance often depends on characteristics of the data. Thus, a thorough characterization of the datasets considered in a benchmark is desirable. Given that benchmarks typically evaluate methods using a range of metrics and datasets, not all of which are relevant to all analysts, flexibility in defining the ranking approach and aggregation of metrics is preferable. Such flexibility, which requires access to various artifacts, is typically not part of published benchmarks today \cite{Sonrel2023-te}. However, a well-structured benchmarking system could accommodate flexible filtering and aggregation of metrics and datasets, and additionally provide access to the code and software stack that the analyst would need to apply the selected method to their own dataset.

@@

Benchmarks can also be of high value to \emph{scientific journals} and \emph{funding agencies}. Well-executed benchmark studies can be highly utilized, influential and cited, and may point to gaps in the current body of methods, guiding future methodological developments; at the same time, in fast-moving fields like bioinformatics, benchmarks have the tendency to rapidly become stale. Journals and funding bodies also have a vested interest in ensuring that published or funded method developments are performed to a high standard, that unnecessary redundancy is reduced, and that results are FAIR (findable, accessible, interoperable and reusable) for maximal benefit to the community \cite{Wilkinson2016-bh}. Thus, quality assurance, neutrality and transparency of benchmarking systems become highly relevant.

@@

Benchmarks are collections of data and source code, ideally executed as workflows within a computing environment. Indeed, recent studies highlight that code is frequently open, reusable and versioned, as are input datasets \cite{Sonrel2023-te, cao2023-jz}. However, \textit{extensibility} of benchmarking studies is generally very low, as is the proportion of benchmarks using a formal workflow system.

Over 350 workflow languages, platforms or systems  \cite{Wratten2021-el, Amstutz2024-qk} and one standard (Common Workflow Language, CWL) exist \cite{Amstutz2016-vo}. Some workflow languages can export workflow definitions to CWL, hence helping computational FAIR guidelines: that is, bundling and processing data together with their metadata, tracking new metadata generation, recording provenance, and being accessible \cite{Goble2020-ps}. For benchmarking, we believe both the workflow layout and provenance can be formally defined (Figure~\ref{fig:formalization}).

@@

Benchmarking as a process deals with tasks other than workflow definition and execution, including: collecting and keeping track of contributions from users; provisioning or being aware of homogeneous and comparable hardware; handling a reproducible and efficient software stack; managing storage and access-control lists; rendering results dashboards and summarizing results; versioning code, workflow runs, and files; and providing extensive documentation from defining scope or contribution guidelines, to transparent logging and crediting contributions and disseminating results.

@@

Parameter choices influence results interpretation, as they can be included in a benchmark to test performance across a wider range of settings, or to find the optimal parameter settings. 
Dataset-specific parameters, e.g., the true number of clusters when evaluating clustering algorithms, are not method-specific. They can also be seen as (dataset-specific) metadata. To differentiate between the effect of the parameter and the methods performances themselves, data-specific parameters should be shared across methods during evaluation.
Similarly, optimizing method-specific parameters for some but not other methods can skew results and should be avoided. 
Aside of optimal and method-specific parameter searches, comparing a method using different settings against other methods can be viewed as a comparing fully distinct methods. This should be clearly communicated in the results. 
For continuous and collaborative benchmarking, parameter optimization and usage should be governed and coordinated between all contributors, ideally ensuring comparability and interpretability of the results.

@@

Most benchmarks could begin decentralized, allowing for concept definition and preliminary results before `going public' (centralization phase). Users need a system that supports building benchmarks both ways, transitioning to centralization as the design matures. However, becoming centralized is not cost-free: it implies gatekeeping (contributors authenticate to the system and get access to computational resources) and adding measures to build trust \textit{via} transparency.

@@

Neutrality is a highly desirable property in a benchmark: it means that efforts have been taken to ensure impartiality or unbiasedness (or at least transparency) of the various choices made (e.g., what datasets to include, how exactly to run methods, how to evaluate methods)\cite{weber2019-el,jelizarow2010-kr}. One could claim that all MDP benchmarks are non-neutral, since the goal is generally to highlight the virtues of a new method. Similarly, ground truth generation and simulations in MDP benchmarks can unconsciously mirror the method capabilities, mechanisms, or strengths, thus biasing evaluation. Furthermore, there are many field experts who may wish to conduct a benchmark (or participate in a collaborative benchmark) on analyses that involve their own tools, where neutrality is then harder to establish. Here,  recent work to mitigate this conflict involves pre-registration and posting the parameters of the study in advance at a registry \cite{Sullivan2019-wh, Olevska2021-if}; initial benchmark formalization could even provide the basis for the formal preregistration. In contrast, BOP benchmarks give, in principle, a higher degree of neutrality, but the challenge to assess neutrality directly remains. Other benchmarking models are possible, from challenges to hackathons, with the hope that a consensus of the masses can further help to establish neutrality. One potential strategy, in the context of the systematic design of benchmarks, is establishing a system where a benchmark definition is parsed into a pre-registration template. Templates exist for simulation studies that could be adapted more generally for benchmarking projects \cite{Siepe2023-tz}.

Finally, there is a tension between preregistration and the concept of `continuous' benchmarking, where new datasets, methods and new metrics are added over time. The logic of continuous benchmarking is that as the problem becomes more understood in the field, it becomes increasingly clear what are good ways to evaluate the performance of methods.

@@

Software plays an important role in the current scientific landscape \cite{Howison2015-rc}, since it is widely used throughout the scientific process, including data collection, simulation, analysis and reporting. Researchers are encouraged to follow FAIR guidelines and reproducibility in their work. Ideally, this would increase benchmark (software) reuse by the wider scientific community. In practice, however, it is sometimes easier to develop new code than to reuse existing software \cite{Trisovic2022-ol}. This leads to the phenomenon of `academic abandonware', where projects are forgotten in code repositories and not maintained after delivering a research output.

Operational costs are given by the expected compute and storage requirements. In the context of benchmarking, methods under evaluation typically have variable computational demands. Benchmark designers could impose resource limits, thereby evaluating the methods under constraints. Storage costs, on the other hand, are more predictable and can be directly influenced by design decisions of the benchmark (e.g., low retention for re-generatable artifacts, `cold storage' for code archives and software, external data stores such as Zenodo for method performance artifacts, CernVM-FS for reusable software).

@@

The UNIX design philosophy \cite{Pike1984-we} is worth embracing to produce software for benchmarking systems that are lightweight, robust, and easily maintainable; ease of use and extensibility are also desirable. Furthermore, to reduce the entry barrier for contributors, the benchmarking system should be capable of running locally with minimal setup. This approach ensures that the system remains relevant and functional even in the absence of centralization. Furthermore, intuitive APIs and sensible defaults \cite{Proctor2018-do} are factors that contribute to highly usable systems. Structured approaches to documentation, including adopting frameworks like diataxis \cite{ProcidaUnknown-vf}, can facilitate adoption and new contributions to a benchmark.

@@

Computational reproducibility comes from controlling the triad of data, code, and environment \cite{Hill2024-gf}. During benchmarking, a workflow runs code transforming inputs into outputs: this execution is affected by the base system (OS, compiler toolchains, libraries) and its configuration. The benchmark definition should control the execution environment.

@@

Even a simple, single-person benchmark comparing methods relies on multiple contributions, from dataset creators and method developers to system software and workflow authors. Hence, data, code and operating system dependencies are all accessible during the benchmarking process. For reproducibility, transparency and to build trust \cite{Laine2007-py}, this accessibility and freedom should be extended for others to run (re-run), copy (including fork), distribute (e.g., snapshot), study (e.g., read how someone is implementing a method), change (bug fix, contribute) and improve (extend) benchmarks. Such freedoms are granted (or restricted) by software and content licenses, as are copyright and authorship and attribution \cite{Kreutzer2014-ua}. Without explicit licenses, even publicly available code is copyrighted by default, and reuse is restricted (to reproduce, distribute, or create derivative works) \cite{Kreutzer2014-ua}. A concise summary of license rights and restrictions, and compatibility across commonly-used open-source licenses can be found in Table~\ref{tab:licenses}. Benchmark contributions should be licensed and associated to scholarly names including email and ORCID (as the canonical identifier), hence allowing seamless reuse, transparency and attribution.

@@

\begin{itemize}
    \item Flexibility vs. constraints:  An unconstrained benchmark grants method contributors greater flexibility but requires more effort from the benchmark maintainer (maintenance, version support). Introducing constraints (such as software dependencies or common data formats) adds initial setup work but can boost maintainability through built-in validation and quality checks.
    \item  Achieving software reproducibility needs discipline and commitment \cite{Lamb2021-tb}. A benchmarker can state the desired degree of replicability and constraints; different tasks require different levels of compliance. 
    \item  Security concerns: decentralized runs lower the participation barrier, but also increase the attack  surface. Sandboxed environments (restricting network access, filesystem access etc) can be provided as mitigation. Additionally, static and dynamic code analysis in the continuous integration process can aid in approving potentially risky submissions.
\end{itemize}

@@

Benchmarking involves applying methods to data. Data types can be diverse and even a single dataset can be stored in multiple formats. A broad distinction can be made according to general vs. specific formats and open vs. proprietary formats. File formats specific to a programming language should be avoided since those formats may change over time, security risks can arise \cite{Huynh2023-pq, Bleih2024-lv}, and in general they are not interoperable. Likewise, proprietary formats should be avoided, since they may change and could require specific (proprietary) software to read. General language-agnostic formats often do not change over time and thus align to FAIR principles \cite{Wilkinson2016-bh}. Generally, using established standards is recommended (e.g., SAM, BED, FASTQ, etc for genomics), since related methods are developed around these formats.

@@

Benchmarking is a cornerstone practice in bioinformatics, impacting how methods are conceived, developed, published and chosen. Benchmarking is driven by the effort of people with the expertise to address (biological and technical) problems. This expertise can be leveraged by current practices of community benchmarking, so scientific advance is eased by joint efforts to catalog open problems, generate (or simulate) appropriate input data, develop methods to address them, and evaluate their performances. Community benchmarking has been proved successful and very productive \cite{Moult2005-ne,Capella-Gutierrez2017-dh,Luecken2024-fk}.

Here we discussed the motivation and process of community benchmarking at several levels, from definition to execution, while highlighting technologies to potentially ease the technical, scientific and contribution challenges.

>>>>

Benchmarking (further defined below) is a critical step in method development. Such studies are typically published as either `methods-development papers' (MDPs) where new methods are compared against existing ones; or `benchmark-only papers' (BOPs), where a set of existing methods are compared in a more neutral way, as defined in \cite{cao2023-jz}. A benchmarking study can be run by a single person or a small group, or it can be organized as a `challenge' or within a hackathon. In this work, we present a perspective on building a robust benchmarking ecosystem in bioinformatics, coming primarily from the methodologist's viewpoint (researchers developing computational tools).  While our focus is bioinformatics, we aim to abstract principles and connect to similar challenges in adjacent fields.

In our discussion, we outline concepts and specifications for an ideal benchmarking system; we examine philosophical issues like gatekeeping and trust; and provide some technical recommendations, such as standardization.

@@

Benchmarks are useful for a broad set of stakeholders. For a \emph{data analyst}, with an aim to perform a specific type of analysis on a specific dataset, benchmarks are helpful tools to find a suitable method for the task. Hence, it is important that benchmarks include datasets similar to the one of interest to the analyst, as method performance often depends on characteristics of the data. Thus, a thorough characterization of the datasets considered in a benchmark is desirable. Given that benchmarks typically evaluate methods using a range of metrics and datasets, not all of which are relevant to all analysts, flexibility in defining the ranking approach and aggregation of metrics is preferable. Such flexibility is typically not part of published benchmarks today \cite{Sonrel2023-te}. However, a well-structured benchmarking system could accommodate flexible filtering and aggregation of metrics and datasets, and additionally provide access to the code and software stack that the analyst would need to apply the selected method to their own dataset.

@@

Benchmarks can also be of high value to \emph{scientific journals} and \emph{funding agencies}. Well-executed benchmark studies can be highly utilized and influential, and may point to gaps in the current body of methods, guiding future methodological developments; at the same time, in fast-moving fields like bioinformatics, benchmarks have the tendency to rapidly become stale. Journals and funding bodies also have a vested interest in ensuring that published or funded method developments are performed to a high standard, that unnecessary redundancy is reduced, and that results are FAIR (findable, accessible, interoperable and reusable) for maximal benefit to the community \cite{Wilkinson2016-bh}. Thus, quality assurance, neutrality and transparency of benchmarking systems become highly relevant.

@@

Benchmarks are collections of data and source code, ideally executed as workflows within a computing environment. Indeed, recent studies highlight that code is frequently open, reusable and versioned, as are input datasets \cite{cao2023-jz, Sonrel2023-te}. However, \textit{extensibility} of benchmarking studies is generally very low, as is the proportion of benchmarks using a formal workflow system.

Over 350 workflow languages, platforms or systems  \cite{Wratten2021-el, Amstutz2024-qk} and one standard (Common Workflow Language, CWL) exist \cite{Amstutz2016-vo}. Some workflow languages can export workflow definitions to CWL, hence helping computational FAIR guidelines, i.e., bundling and processing data together with their metadata, tracking new metadata generation, recording provenance, and being accessible \cite{Goble2020-ps}. For benchmarking, we believe both the workflow layout and provenance can be formally defined (Figure~\ref{fig:formalization}).

@@

Benchmarking as a process deals with tasks other than workflow definition and execution, including: collecting and keeping track of contributions from users; provisioning appropriate hardware; handling a reproducible and efficient software stack; managing storage and access-control lists; rendering results dashboards and summarizing results; versioning code, workflow runs, and files; and providing extensive documentation from defining scope or contribution guidelines, to transparent logging and crediting contributions and disseminating results.

@@

Parameter choices influence results interpretation, as they can be included in a benchmark to test performance across a wider range of settings, or to find the optimal parameter settings. 
Dataset-specific parameters, e.g., the true number of clusters when evaluating clustering algorithms, are not method-specific; they can be seen as (dataset-specific) metadata. To differentiate between the effect of the parameter and the methods performances themselves, data-specific parameters should be shared across methods during evaluation.
Similarly, optimizing method-specific parameters for some but not other methods can skew results and should be avoided. 
Aside of optimal and method-specific parameter searches, comparing a method under different settings against other methods can be viewed as comparing fully distinct methods, which should be clearly communicated in the results. 
For continuous and collaborative benchmarking, parameter optimization and usage should be governed and coordinated between all contributors, ideally ensuring comparability and interpretability of the results.

@@

Most benchmarks could begin decentralized, allowing for concept definition and preliminary results before `going public' (centralization phase). Users need a system that supports building benchmarks both ways, transitioning to centralization as the design matures. However, becoming centralized is not cost-free: it implies gatekeeping (contributors authenticate to the system and get access to computational resources) and requires additional measures that build trust (e.g., transparency).

@@

Neutrality is a highly desirable property in a benchmark: it means that efforts have been taken to ensure impartiality or unbiasedness (or at least transparency) of the various choices made (e.g., what datasets to include, how exactly to run methods, how to evaluate methods) \cite{Weber2019-el, Jelizarow2010-kr}. One could claim that all MDP benchmarks are non-neutral, since the goal is generally to highlight the virtues of a new method. Similarly, ground truth generation and simulations in MDP benchmarks can unconsciously mirror the method capabilities, mechanisms, or strengths, thus biasing evaluation. Furthermore, there are many field experts who may wish to conduct a benchmark (or participate in a collaborative benchmark) on analyses that involve their own tools, where neutrality is then harder to establish. Here,  recent work to mitigate this conflict involves pre-registration and posting the parameters of the study in advance at a registry \cite{Sullivan2019-wh, Olevska2021-if}. In contrast, BOP benchmarks give, in principle, a higher degree of neutrality, but the challenge to assess neutrality directly remains. Other benchmarking models are possible, from challenges to hackathons, with the hope that a consensus of the masses can further help to establish neutrality. One potential strategy, in the context of the systematic design of benchmarks, is establishing a system where a benchmark definition is parsed into a pre-registration template. Templates exist for simulation studies that could be adapted more generally for benchmarking projects \cite{Siepe2023-tz}.

Finally, there is a tension between preregistration and the concept of `continuous' benchmarking, where new datasets, methods and new metrics are added over time. The logic of continuous benchmarking is that as the computational task becomes more understood, it becomes increasingly clear how to best evaluate the performance of methods.

@@

Software plays an important role in the current scientific landscape \cite{Howison2015-rc}, since it is widely used throughout the scientific process, including data collection, simulation, analysis and reporting. Even if open sourced, however, it is sometimes easier to develop new code than to reuse existing software \cite{Trisovic2022-ol}. This leads to the phenomenon of `academic abandonware', where projects are forgotten in code repositories and not maintained after delivering a research output.

Operational costs of benchmarks are given by the expected compute and storage requirements. In the context of benchmarking, methods under evaluation typically have variable computational demands. Benchmark designers could impose resource limits, thereby evaluating the methods under constraints. Storage costs, on the other hand, are more predictable and can be directly influenced by design decisions of the benchmark (e.g., low retention for re-generatable artifacts, `cold storage' for code archives and software, external data stores such as Zenodo for method performance artifacts, CernVM-FS for reusable software).

@@

The UNIX design philosophy \cite{Pike1984-we} is worth embracing to produce software for benchmarking systems that are lightweight, robust, and easily maintainable; ease of use and extensibility are also desirable. Furthermore, to reduce the entry barrier for contributors, the benchmarking system should be capable of running locally with minimal setup. This approach ensures that the system remains relevant and functional even in the absence of centralization. Furthermore, intuitive APIs and sensible defaults \cite{Proctor2018-do} are factors that contribute to highly usable systems. Structured approaches to documentation, including frameworks such as diataxis \cite{ProcidaUnknown-vf}, can facilitate adoption and new contributions to a benchmark.

@@

Computational reproducibility comes from controlling the triad of data, code, and environment \cite{Hill2024-gf}. During benchmarking, a workflow runs code transforming inputs into outputs; execution is affected by the base system (OS, compiler toolchains, libraries) and its configuration. The benchmark definition should control the execution environment.

@@

Even a simple, single-person benchmark comparing methods relies on multiple contributions, from dataset creators and method developers to system software and workflow authors. Hence, data, code and operating system dependencies are all accessible during the benchmarking process. For reproducibility, transparency and to build trust \cite{Laine2007-py}, this accessibility and freedom should be extended to others to run (re-run), copy (including fork), distribute (e.g., snapshot), study (e.g., read how someone is implementing a method), change (bug fix, contribute) and improve (extend) benchmarks. Such freedoms are granted (or restricted) by software and content licenses, as are copyright and authorship and attribution \cite{Kreutzer2014-ua}. Without explicit licenses, even publicly available code is copyrighted by default, and reuse is restricted (to reproduce, distribute, or create derivative works) \cite{Kreutzer2014-ua}. A concise summary of license rights and restrictions, and compatibility across commonly-used open-source licenses can be found in Table~\ref{tab:licenses}. Benchmark contributions should be licensed and associated to scholarly names including email and ORCID (as the canonical identifier), hence allowing seamless reuse, transparency and attribution.

@@

\begin{itemize}
    \item Flexibility vs. constraints:  An unconstrained benchmark grants method contributors greater flexibility but requires more effort from the benchmarker (maintenance, version support). Introducing constraints (such as software dependencies or common data formats) adds initial setup work but can boost maintainability through built-in validation and quality checks.
    \item  Achieving software reproducibility needs discipline and commitment \cite{Lamb2021-tb}. A benchmarker can state the desired degree of replicability and constraints. 
    \item  Security concerns: decentralized runs lower the participation barrier, but also increase the attack  surface. Sandboxed environments (restricting network access, filesystem access etc) can be provided as mitigation. Additionally, static and dynamic code analysis in the continuous integration process can aid in approving potentially risky submissions.
\end{itemize}

@@

Benchmarking involves applying methods to data. Data types can be diverse and even a single dataset can be stored in multiple formats. A broad distinction can be made according to general vs. specific formats and open vs. proprietary formats. File formats specific to a programming language should be avoided since those formats may change over time, security risks can arise \cite{Huynh2023-pq, Bleih2024-lv}, and they are not always interoperable. Likewise, proprietary formats should be avoided, since they may change and could require specific (proprietary) software to read. General language-agnostic formats often do not change over time and thus align to FAIR principles \cite{Wilkinson2016-bh}. Generally, using established standards is recommended (e.g., SAM, BED, FASTQ, etc for genomics), since related methods are developed around these formats.

@@

Benchmarking is a cornerstone practice in bioinformatics, impacting how methods are conceived, developed, published and chosen. Benchmarking is driven by the effort of people with the expertise to address biological and technical problems. This expertise can be leveraged by current practices of community benchmarking, so scientific advances are eased by joint efforts to catalog open problems, generate (or simulate) appropriate input data, develop methods to address them, and evaluate their performances. Community benchmarking has been proven successful and  productive \cite{Moult2005-ne,Capella-Gutierrez2017-dh,Luecken2024-fk}.

Here, we discussed the motivation and process of community benchmarking at several levels, from definition to execution, while highlighting strategies and technologies to potentially ease the technical, scientific and contribution challenges.
